\ttitle

### 2013-10-07 Chapters/Chapter3/one_class_svm.tex
Wip

%--------------------------------------------
\subsection{Support Vector Machine}\label{subsec:svm}
We will first discuss the traditional two-class \gls{svm} before we consider the one-class variant, as introduced by Cortes and Vapnik in \cite{cortes1995support}.
Consider a data set $\Omega = \{ (x_1, y_1),\allowbreak (x_2, y_2), \dots , (x_n, y_n) \}$; points $x_i \in \mathbb{R}^d$ in a (for instance two-dimensional) space where $x_i$ is the $i$-th input data point and $y_i \in \{-1, 1\}$ is the $i$-th output pattern, indicating the class membership.

A \gls{svm} can create a boundary between linear-separable data points, making it a non-probabilistic binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in Figure \ref{fig:kernel_mapping}.
This function maps the input data from space $I$ to a higher dimensional space $F$.
The \gls{svm} can create a linear separating hyperplane in the space $F$ that separates the data points from the two classes.
When the hyperplane is projected to the (lower) original input space $I$ it creates a non-linear separating curve.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in section \ref{subsec:kernel_trick}.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{./Figures/chapter3/svm_separating_plane_with_margin.png}
  \caption[\gls{svm} and the separating hyperplane]{Illustration of the separating hyperplane of a \gls{svm}. Here $w$ is the normal vector for the separating hyperplane and the distance between the two margins is $\frac{2}{\lVert{w}\rVert}$. *** Add wikipedia reference. How to do? ***}
  \label{fig:svm_hyperplane}
\end{figure}

@@

This minimization problem can be solved (using \gls{qp}) and transformed to its Lagrange dual formulation.
To do so, the Lagrange multipliers $a_i >= 0$ are introduced and the decision function becomes:
\begin{equation}\label{eq:svm_lagrange}
  f(x) = \operatorname{sgn}( \sum_{i=1}^n \alpha_i y_i K(x, x_i) + b),
\end{equation}
where $K(x, x_i) = \phi(x)^T\phi(x_i)$ (which is further discussed in section \ref{subsec:kernel_trick}).
Here every data point $i$ for which $a_i > 0$ is weighted in the decision function and thus ``supports'' the classification machine: hence the name ``\acrlong{svm}''.
Since it is shown that under certain circumstances \gls{svm}s show an equality to sparse representations \cite{girosi1998equivalence}, there will often be relatively few Lagrange multipliers with a non-zero value.

@@

\subsection{Kernel trick}\label{subsec:kernel_trick}
In the previous section, \ref{subsec:svm}, the mapping function $\phi(x)$ and the kernel function $K$ were briefly mentioned.
The decision function in equation \ref{eq:svm_lagrange} only relies on the dot-products of mapped data points in the feature space $F$ (\ie all pairwise distances between the data points in that space).
It shows \cite{flach2012machine} that as long as any function has the same result, without an explicit mapping to the higher dimension $F$, the dot-products can be substituted by the kernel function $K$.
This is known as the \emph{kernel trick} and gives the \gls{svm} the ability to create non-linear decision function without high computational complexity.
This mapping is illustrated in Figure \ref{fig:kernel_mapping}.
Here the non-linear separating boundary in the input space $I$ is mapped, via $\phi$, to a linear boundary in the feature space $F$.

The kernel function $K$ can have different forms, such as linear, polynomial and sigmoidal but the most used (and flexible) form is the Gaussian \gls{rbf}, in which the feature space $F$ is a Hilbert Space of infinite dimensions *** ref needed ***:
\begin{equation}
  K(x, x') = \operatorname{exp} \left( - \frac{ \lVert x - x' \rVert ^2}{2 \sigma^2 } \right),
\end{equation}
where $\sigma \in R$ is a kernel parameter and $\lVert x - x' \rVert$ is the dissimilarity measure expressed in Euclidean distance.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth]{./Figures/chapter3/svm_kernel_mapping.png}
  \caption[Kernel mapping]{The non-linear boundary in the input space $I$ (left) is transformed to a linear boundary in the feature space $F$ (right) by mapping the data points with the function $\phi$. The kernel trick uses a function $K$ which performs an implicit mapping. *** Add wikipedia reference. How to do? ***}
  \label{fig:kernel_mapping}
\end{figure}

@@

The constructed decision function $f$ resembles the function discussed in Section \ref{subsec:svm}.
It returns the value $+1$ in a (possibly small) region capturing most of the data points, and $-1$ elsewhere.
The method maps the data points from input space $I$ to a feature space $F$ (following classical \gls{svm}s).
In that space $F$ it separates the data points with maximal margin from the origin, with a hyperplane.
For a new data points $x$, the function value $f(x)$ determines wheter the data point is part of the distribution (\ie the value is $+1$) or a novelty (\ie the value is $-1$).
The hyperplane is represented by $g(x) = w \cdot \phi(x) + \rho = 0$ and the decision function is $f(x) = \operatorname{sgn}(g(x))$.
This hyperplane and the separation from the origin is illustrated in Figure \ref{fig:nu-svm}.

@@

This method creates a \emph{hyperplane}, characterized by $w$ and $\rho$, that separates the data with maximal margin from the origin in the feature space $F$.
In the following section we will discuss an alternative method, which uses an circumscribing \emph{hypersphere} to characterize the data (and the region of distribution $P$ where the density (or: support) lives).

@@

\subsection{\acrlong{svdd}}\label{subsec:svdd}
The method introduced by Tax and Duin \cite{tax2004support}, known as \acrlong{svdd}, takes as spherical instead of planar approach.
The boundary, created in feature space $F$, forms a hypersphere around the (high density region of the) data.
The volumne of this hypersphere is minimized to get the smallest enclosing boundary.
By allowing outliers using slacks variables, in the same manner as classical \gls{svm} and \gls{nu-svm}, a soft margin is constructed.

The constructed hypersphere is characterized by a center $\mathbf{a}$ and a radius $R > 0$ as distance from the center to (any data point that is a \gls{sv} on) the boundary, for which the volume, and thus the radius $R$, will be minimized.

@@

>>>>

%--------------------------------------------
\subsection{Support Vector Machine}\label{subsec:svm}
We will first discuss the traditional two-class \gls{svm} before we consider the one-class variant, as introduced by Cortes and Vapnik in \cite{cortes1995support}.
Consider a data set $\Omega = \{ (x_1, y_1),\allowbreak (x_2, y_2), \dots , (x_n, y_n) \}$; points $x_i \in \mathbb{R}^d$ in a (for instance two-dimensional) space where $x_i$ is the $\mathcal{I}$-th input data point and $y_i \in \{-1, 1\}$ is the $\mathcal{I}$-th output pattern, indicating the class membership.

A \gls{svm} can create a boundary between linear-separable data points, making it a non-probabilistic binary linear classifier.
More flexible non-linear boundaries can be obtained by the use of a non-linear function $\phi(x)$, as illustrated in Figure \ref{fig:kernel_mapping}.
This function maps the input data from space $\mathcal{I}$ to a higher dimensional space $\mathcal{F}$.
The \gls{svm} can create a linear separating hyperplane in the space $\mathcal{F}$ that separates the data points from the two classes.
When the hyperplane is projected to the (lower) original input space $\mathcal{I}$ it creates a non-linear separating curve.
The mapping and projection of data points can be efficient (and implicit) performed by using the kernel trick, which is discussed in section \ref{subsec:kernel_trick}.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{./Figures/chapter3/svm_separating_plane_with_margin.png}
  \caption[\gls{svm} and the separating hyperplane]{Illustration of the separating hyperplane of a \gls{svm}. Here $w$ is the normal vector for the separating hyperplane and the distance between the two margins is $\frac{2}{\lVert{w}\rVert}$. Image from Wikipedia.org\protect\footnote}
  \label{fig:svm_hyperplane}
\end{figure}
\footnotetext{\url{http://en.wikipedia.org/wiki/File:Svm_max_sep_hyperplane_with_margin.png}}

@@

This minimization problem can be solved (using \gls{qp}) and transformed to its Lagrange dual formulation.
To do so, the Lagrange multipliers $a_i >= 0$ are introduced and the decision function becomes:
\begin{equation}\label{eq:svm_lagrange}
  f(x) = \operatorname{sgn}( \sum_{i=1}^n \alpha_i y_i K(x, x_i) + b),
\end{equation}
where $K(x, x_i) = \phi(x)^T\phi(x_i)$ (which is further discussed in section \ref{subsec:kernel_trick}).
Here every data point $\mathcal{I}$ for which $a_i > 0$ is weighted in the decision function and thus ``supports'' the classification machine: hence the name ``\acrlong{svm}''.
Since it is shown that under certain circumstances \gls{svm}s show an equality to sparse representations \cite{girosi1998equivalence}, there will often be relatively few Lagrange multipliers with a non-zero value.

@@

\subsection{Kernel trick}\label{subsec:kernel_trick}
In the previous section, \ref{subsec:svm}, the mapping function $\phi(x)$ and the kernel function $K$ were briefly mentioned.
The decision function in equation \ref{eq:svm_lagrange} only relies on the dot-products of mapped data points in the feature space $\mathcal{F}$ (\ie all pairwise distances between the data points in that space).
It shows \cite{flach2012machine} that as long as any function has the same result, without an explicit mapping to the higher dimension $\mathcal{F}$, the dot-products can be substituted by the kernel function $K$.
This is known as the \emph{kernel trick} and gives the \gls{svm} the ability to create non-linear decision function without high computational complexity.
This mapping is illustrated in Figure \ref{fig:kernel_mapping}.
Here the non-linear separating boundary in the input space $\mathcal{I}$ is mapped, via $\phi$, to a linear boundary in the feature space $\mathcal{F}$.

The kernel function $K$ can have different forms, such as linear, polynomial and sigmoidal but the most used (and flexible) form is the Gaussian \gls{rbf}, in which the feature space $\mathcal{F}$ is a Hilbert Space of infinite dimensions *** ref needed ***:
\begin{equation}
  K(x, x') = \operatorname{exp} \left( - \frac{ \lVert x - x' \rVert ^2}{2 \sigma^2 } \right),
\end{equation}
where $\sigma \in R$ is a kernel parameter and $\lVert x - x' \rVert$ is the dissimilarity measure expressed in Euclidean distance.

\begin{figure}
\centering
  \includegraphics[width=0.8\textwidth]{./Figures/chapter3/svm_kernel_mapping.png}
  \caption[Kernel mapping]{The non-linear boundary in the input space $\mathcal{I}$ (left) is transformed to a linear boundary in the feature space $\mathcal{F}$ (right) by mapping the data points with the function $\phi$. The kernel trick uses a function $K$ which performs an implicit mapping. Image from Wikipedia.org\footnotemark}
  \label{fig:kernel_mapping}
\end{figure}
\footnotetext{\url{http://en.wikipedia.org/wiki/File:Kernel_Machine.png}}

@@

The constructed decision function $\mathcal{F}$ resembles the function discussed in Section \ref{subsec:svm}.
It returns the value $+1$ in a (possibly small) region capturing most of the data points, and $-1$ elsewhere.
The method maps the data points from input space $\mathcal{I}$ to a feature space $\mathcal{F}$ (following classical \gls{svm}s).
In that space $\mathcal{F}$ it separates the data points with maximal margin from the origin, with a hyperplane.
For a new data points $x$, the function value $f(x)$ determines wheter the data point is part of the distribution (\ie the value is $+1$) or a novelty (\ie the value is $-1$).
The hyperplane is represented by $g(x) = w \cdot \phi(x) + \rho = 0$ and the decision function is $f(x) = \operatorname{sgn}(g(x))$.
This hyperplane and the separation from the origin is illustrated in Figure \ref{fig:nu-svm}.

@@

This method creates a \emph{hyperplane}, characterized by $w$ and $\rho$, that separates the data with maximal margin from the origin in the feature space $\mathcal{F}$.
In the following section we will discuss an alternative method, which uses an circumscribing \emph{hypersphere} to characterize the data (and the region of distribution $P$ where the density (or: support) lives).

@@

\subsection{\acrlong{svdd}}\label{subsec:svdd}
The method introduced by Tax and Duin \cite{tax2004support}, known as \acrlong{svdd}, takes as spherical instead of planar approach.
The boundary, created in feature space $\mathcal{F}$, forms a hypersphere around the (high density region of the) data.
The volume of this hypersphere is minimized to get the smallest enclosing boundary.
By allowing outliers using slacks variables, in the same manner as classical \gls{svm} and \gls{nu-svm}, a soft margin is constructed.

The constructed hypersphere is characterized by a center $\mathbf{a}$ and a radius $R > 0$ as distance from the center to (any data point that is a \gls{sv} on) the boundary, for which the volume, and thus the radius $R$, will be minimized.
The center $\mathbf{a}$ is a linear combination of of the support vectors.
Like the classical \gls{svm} and \gls{svdd} it can be required that all the distances from the data points $x_i$ to the center $\mathbf{a}$ are strict less then $R$ (or equivalent measure, to create a hard margin) or soft margins are allowed by using slack variables $\xi_i$.
In the case of a soft margin, the penalty is determined by $C$ and the minimization is expressed as Equation \ref{eq:svdd_objective}.
This principle is illustrated in the right image or Figure \ref{fig:nu-vs-svdd}.
Instead of a separating hyperplane, constructed by \gls{nu-svm} and illustrated on the left of the Figure, the \gls{svdd} creates a hypersphere (in the illustration a cricle) around the data points.
By using kernel functions (\eg the \gls{rbf}) the hyperspheres in the high dimensional feature space $\mathcal{F}$ corresponds to a flexible and tight enclosing boundary in input space $\mathcal{I}$, illustrated in Figure \ref{fig:svdd-boundary}.

@@

The dual formulation, by using Lagrange multipliers $\alpha_i$, can be used to test a new data point $z$ and is the following:

\begin{equation}
  \lVert z - \mathbf{x} \rVert ^2 = \sum_{i=1}^n \alpha_i \operatorname{exp} \left( \frac{- \lVert z - x_i \rVert ^ 2}{\sigma^2} \right) \geq -R^2/2 + C_R
\end{equation}

\subsection{Kernel Function}\label{subsec:kernel_function}

\begin{itemize}
  \item Explain RBF kernel
  \item Explain RKHS? (Hilbert Space)
\end{itemize}

\begin{figure}
  \centering
    \includegraphics[width=0.5\textwidth,keepaspectratio]{./Figures/chapter3/svdd-boundary.pdf}
  \caption[\gls{svdd} boundary]{The \gls{svdd} method trained on a banana-shaped data set with different sigma-values for the \gls{rbf} kernel. Solid circles are support vectors, the dashed line is the boundary. Image by Tax \cite{tax2001one}.}
  \label{fig:svdd-boundary}
\end{figure}
